---
title: "ST5209/X Assignment 4"
format: pdf
editor: visual
author: "Jiao Dian"
execute:
  warning: false
---

## Submission

1.  Render the document to get a .pdf printout.
2.  Submit both the .qmd and .pdf files to Canvas.

```{r}
#| message: FALSE
#| echo: FALSE
#| output: false
Sys.setlocale("LC_ALL", "English")
options(repos = c(CRAN = "https://cloud.r-project.org"))
library(tidyverse)
library(fpp3)
```

## 1. AR polynomial

Consider the AR(2) model $$
X_t = 4 + 0.5 X_{t-1} - 0.25 X_{t-2} + W_t.
$$ {#eq-ar2}

a.  **Autoregressive polynomial:**

The AR polynomial is obtained by writing the model as $X_t - 0.5X_{t-1} + 0.25X_{t-2} = 4 + W_t$, giving $$
\phi(z) = 1 - 0.5z + 0.25z^2.
$$

b.  **Roots:**

We solve $\phi(z) = 0$, i.e. $0.25z^2 - 0.5z + 1 = 0$. Multiplying by 4: $z^2 - 2z + 4 = 0$.

$$
z = \frac{2 \pm \sqrt{4 - 16}}{2} = \frac{2 \pm \sqrt{-12}}{2} = 1 \pm i\sqrt{3}.
$$

```{r}
polyroot(c(1, -0.5, 0.25))
```

The roots are $z_1 = 1 + i\sqrt{3}$ and $z_2 = 1 - i\sqrt{3}$.

c.  **Causality:**

The modulus of each root is $|z| = \sqrt{1^2 + (\sqrt{3})^2} = \sqrt{4} = 2 > 1$. Since both roots lie strictly outside the unit circle, the model is **causal**.

d.  **Linear process coefficients $\psi_0, \psi_1, \psi_2, \psi_3$:**

Following the strategy in `linear_process_coefs.qmd`, we use the relation $\phi(z)\sum_{j=0}^\infty \psi_j z^j = 1$. Writing $\phi_0 = 1$, $\phi_1 = -0.5$, $\phi_2 = 0.25$, the recursive formula is $\sum_{j=0}^{2} \phi_j \psi_{k-j} = 0$ for $k \geq 1$, i.e. $\psi_k = 0.5\psi_{k-1} - 0.25\psi_{k-2}$, with $\psi_0 = 1$ and $\psi_j = 0$ for $j < 0$.

-   $\psi_0 = 1$
-   $k=1$: $\psi_1 + (-0.5)(1) = 0 \Rightarrow \psi_1 = 0.5$
-   $k=2$: $\psi_2 + (-0.5)(0.5) + (0.25)(1) = 0 \Rightarrow \psi_2 = 0.25 - 0.25 = 0$
-   $k=3$: $\psi_3 + (-0.5)(0) + (0.25)(0.5) = 0 \Rightarrow \psi_3 = -0.125$

Verification using R:

```{r}
ARMAtoMA(ar = c(0.5, -0.25), lag.max = 3)
```

So $\psi_0 = 1$, $\psi_1 = 0.5$, $\psi_2 = 0$, $\psi_3 = -0.125$.

The mean is $\mu = \frac{4}{1 - 0.5 + 0.25} = \frac{4}{0.75} = \frac{16}{3}$.

## 2. Likelihood

Consider the AR(1) model $$
X_t = 1 - 0.6X_{t-1} + W_t,
$$ where $W_t \sim WN(0, 0.25)$ is Gaussian white noise. Observations: $x_1 = 0.2, x_2 = -0.3, x_3 = 0.4$.

a.  **Mean of $X_t$:**

Writing the model as $X_t = c + \phi X_{t-1} + W_t$ with $c = 1$ and $\phi = -0.6$, the mean is $$
\mu = \frac{c}{1 - \phi} = \frac{1}{1 - (-0.6)} = \frac{1}{1.6} = 0.625.
$$

b.  **Conditional likelihood given $x_1$:**

Conditioning on $x_1$, the conditional likelihood is $L = p(x_2|x_1) \cdot p(x_3|x_2)$. Since $X_t | X_{t-1} = x_{t-1} \sim N(1 - 0.6x_{t-1},\; 0.25)$, each factor is a Gaussian density.

The conditional means are:

-   $\mu_2 = 1 - 0.6(0.2) = 0.88$
-   $\mu_3 = 1 - 0.6(-0.3) = 1.18$

The residuals are:

-   $e_2 = x_2 - \mu_2 = -0.3 - 0.88 = -1.18$
-   $e_3 = x_3 - \mu_3 = 0.4 - 1.18 = -0.78$

The conditional likelihood is: $$
L = \prod_{t=2}^{3} \frac{1}{\sqrt{2\pi \cdot 0.25}} \exp\left(-\frac{e_t^2}{2 \cdot 0.25}\right)
= \frac{1}{2\pi \cdot 0.25} \exp\left(-\frac{(-1.18)^2 + (-0.78)^2}{0.5}\right)
$$

```{r}
e2 <- -0.3 - (1 - 0.6 * 0.2)
e3 <- 0.4 - (1 - 0.6 * (-0.3))
sigma2 <- 0.25

cat("e2 =", e2, "\n")
cat("e3 =", e3, "\n")
cat("e2^2 + e3^2 =", e2^2 + e3^2, "\n")

L <- prod(dnorm(c(-0.3, 0.4), mean = c(1 - 0.6 * 0.2, 1 - 0.6 * (-0.3)), sd = sqrt(sigma2)))
cat("Conditional likelihood:", L, "\n")
```

The sum of squared residuals is $(-1.18)^2 + (-0.78)^2 = 1.3924 + 0.6084 = 2.0008$, so the conditional likelihood evaluates to approximately $0.0116$. This small value reflects the fact that the observed data points are relatively far from their conditional means, suggesting a poor fit for these particular observations.

## 3. Reversibility

a.  Simulate a sample trajectory of length $n = 500$ from the AR(2) model in Problem 1.

```{r}
set.seed(5209)
xt <- arima.sim(model = list(ar = c(0.5, -0.25)), n = 500)
```

b.  Reverse the time index.

```{r}
xt_rev <- rev(as.numeric(xt))
```

c.  Fit an AR(2) model to the reversed time series.

```{r}
rev_tsibble <- tsibble(
  t = 1:500,
  X = xt_rev,
  index = t
)

rev_fit <- rev_tsibble |>
  model(AR(X ~ order(2)))
```

d.  Inspect the model parameters.

```{r}
rev_fit |> tidy()
```

The fitted AR(2) coefficients are approximately $\hat\phi_1 \approx 0.499$ and $\hat\phi_2 \approx -0.239$, which are close to the true values $\phi_1 = 0.5$ and $\phi_2 = -0.25$ from @eq-ar2. This is because of the **reversibility** property: for a causal stationary process, the autocovariance function $\gamma(h)$ is symmetric ($\gamma(h) = \gamma(-h)$), so the Yule-Walker equations are identical whether we read the time series forward or backward. The small discrepancies from the true values are due to finite-sample estimation variability.

e.  Forecast with $h = 10$.

```{r}
rev_fit |>
  forecast(h = 10)
```

```{r}
rev_fit |>
  forecast(h = 10) |>
  autoplot(rev_tsibble |> filter(t > 450)) +
  labs(title = "Forecast on reversed time series")
```

Since we reversed the time series, forecasting forward on the reversed series corresponds to **backcasting** — predicting values that would have occurred before the start of the original time series. In the original time series, the observations are indexed $t = 1, 2, \ldots, 500$. After reversal, time $t = 1$ in the reversed series corresponds to $t = 500$ in the original, so forecasting at $t = 501, \ldots, 510$ in the reversed series corresponds to predicting $X_0, X_{-1}, \ldots, X_{-9}$ in the original. The forecasts converge toward the process mean $\mu = 0$ as the forecast horizon increases, which is expected since `arima.sim()` generates a zero-mean AR(2) process (without the constant term 4) and the model is causal and stationary.

## 4. Yule-Walker

a.  **Yule-Walker equations** for the AR(2) model $X_t = 0.5X_{t-1} - 0.25X_{t-2} + W_t$ (ignoring the constant since it doesn't affect autocovariances):

Multiplying both sides by $X_{t-h}$ and taking expectations:

-   $h = 0$: $\gamma(0) = 0.5\gamma(1) - 0.25\gamma(2) + \sigma^2$
-   $h = 1$: $\gamma(1) = 0.5\gamma(0) - 0.25\gamma(1)$
-   $h = 2$: $\gamma(2) = 0.5\gamma(1) - 0.25\gamma(0)$

b.  **Matrix form.** Rearranging: $$
\left[\begin{matrix}
1 & -0.5 & 0.25 \\
-0.5 & 1.25 & 0 \\
0.25 & -0.5 & 1
\end{matrix}\right]
\left[\begin{matrix}
\gamma(0) \\ \gamma(1) \\ \gamma(2)
\end{matrix}\right]
= \left[\begin{matrix}
\sigma^2 \\ 0 \\ 0
\end{matrix}\right].
$$

Row derivations:

-   Row 1: $\gamma(0) - 0.5\gamma(1) + 0.25\gamma(2) = \sigma^2$
-   Row 2: $-0.5\gamma(0) + (1 + 0.25)\gamma(1) + 0\cdot\gamma(2) = 0$ (from $\gamma(1) = 0.5\gamma(0) - 0.25\gamma(1)$)
-   Row 3: $0.25\gamma(0) - 0.5\gamma(1) + \gamma(2) = 0$ (from $\gamma(2) = 0.5\gamma(1) - 0.25\gamma(0)$)

c.  **Solve numerically** (assuming $\sigma^2 = 1$):

```{r}
A <- matrix(c(1, -0.5, 0.25,
              -0.5, 1.25, 0,
              0.25, -0.5, 1), nrow = 3, byrow = TRUE)
gamma_vals <- solve(A, c(1, 0, 0))
names(gamma_vals) <- c("gamma(0)", "gamma(1)", "gamma(2)")
gamma_vals
```

We obtain $\gamma(0) \approx 1.270$, $\gamma(1) \approx 0.508$, and $\gamma(2) \approx -0.063$. Note that $\gamma(2) < 0$, which is consistent with the negative AR(2) coefficient $\phi_2 = -0.25$.

d.  **Top left entry of $\Gamma_2^{-1}$:**

For a $2 \times 2$ matrix $\Gamma_2 = \begin{bmatrix} \gamma(0) & \gamma(1) \\ \gamma(1) & \gamma(0) \end{bmatrix}$, the inverse is $\frac{1}{\gamma(0)^2 - \gamma(1)^2}\begin{bmatrix} \gamma(0) & -\gamma(1) \\ -\gamma(1) & \gamma(0) \end{bmatrix}$.

The top left entry is: $$
[\Gamma_2^{-1}]_{1,1} = \frac{\gamma(0)}{\gamma(0)^2 - \gamma(1)^2}.
$$

```{r}
g0 <- gamma_vals[1]
g1 <- gamma_vals[2]
top_left_inv <- g0 / (g0^2 - g1^2)
cat("Top left entry of Gamma_2 inverse:", top_left_inv, "\n")
```

The top left entry is $\frac{\gamma(0)}{\gamma(0)^2 - \gamma(1)^2} \approx \frac{1.270}{1.270^2 - 0.508^2} = \frac{1.270}{1.354} \approx 0.9375$.

Verification using `solve()`:

```{r}
Gamma2 <- matrix(c(g0, g1, g1, g0), nrow = 2)
solve(Gamma2)
```

The top left entry of $\Gamma_2^{-1}$ is indeed $0.9375$, confirming the formula.

e.  **95% confidence interval for $\phi_1$.**

The asymptotic distribution of the Yule-Walker estimator is $\hat\phi \approx N(\phi, \frac{\sigma^2}{n}\Gamma_2^{-1})$. Using $n = 500$ from Problem 3 and the fitted parameters:

```{r}
n <- 500
sigma2_hat <- 1  # white noise variance
se_phi1 <- sqrt(sigma2_hat * top_left_inv / n)
cat("Standard error of phi1_hat:", round(se_phi1, 4), "\n")

phi1_hat <- rev_fit |> tidy() |> filter(term == "ar1") |> pull(estimate)
ci_lower <- phi1_hat - 1.96 * se_phi1
ci_upper <- phi1_hat + 1.96 * se_phi1
cat("phi1_hat:", round(phi1_hat, 4), "\n")
cat("95% CI for phi_1: [", round(ci_lower, 4), ",", round(ci_upper, 4), "]\n")
```

The standard error is $\sqrt{\sigma^2 [\Gamma_2^{-1}]_{1,1} / n} = \sqrt{0.9375/500} \approx 0.0433$. Using the estimated $\hat\phi_1 \approx 0.499$ from Problem 3d, the 95% confidence interval is approximately $[0.414, 0.584]$. This interval contains the true value $\phi_1 = 0.5$, as expected.

## 5. Recursive forecasting for ARMA(1, 1)

Consider the ARMA(1,1) model $X_t - 0.5X_{t-1} = W_t + 0.5W_{t-1}$.

```{r}
set.seed(5209)
n <- 50
wn <- rnorm(n)
xt <- arima.sim(model = list(ar = 0.5, ma = 0.5), innov = wn, n = n)
```

a.  **Truncated white noise estimation.** From the model, $W_t = X_t - 0.5X_{t-1} - 0.5W_{t-1}$. Setting $W_0 = 0$:

```{r}
wn_hat <- rep(0, n)
wn_hat[[1]] <- xt[[1]]
for (i in 2:n) {
  wn_hat[[i]] <- xt[[i]] - 0.5 * xt[[i - 1]] - 0.5 * wn_hat[[i - 1]]
}
```

b.  **Log absolute difference plot:**

```{r}
tibble(t = 1:n,
       log_diff = log(abs(wn - wn_hat))) |>
  ggplot(aes(x = t, y = log_diff)) +
  geom_line() +
  labs(x = "Time", y = "log|W_t - W_hat_t|",
       title = "Log absolute difference between true and estimated white noise")
```

c.  From the plot, the log absolute difference decreases approximately **linearly** over time, which means the truncation error $|W_t - \hat{W}_t|$ decays **exponentially** fast. The rate of decay is governed by the MA coefficient $\theta = 0.5$: since $|\theta| < 1$, the initial approximation error ($W_0 = 0$) is multiplied by $(-\theta)^t = (-0.5)^t$ at each step, leading to exponential convergence.

This has an important consequence for truncated forecasts: even with the crude initialization $\hat{W}_0 = 0$, the estimated white noise values $\hat{W}_t$ become virtually indistinguishable from the true $W_t$ after about 20--30 time steps. Therefore, the truncated forecast is an excellent approximation to the exact forecast for moderate to large $n$.

d.  **Truncated forecast for $X_{53}$** (i.e. $h = 3$ steps ahead from $n = 50$).

For the ARMA(1,1) model, the recursive forecast from time $n$ is:

-   $h = 1$: $\hat{X}_{51|50} = 0.5X_{50} + 0.5\hat{W}_{50}$
-   $h = 2$: $\hat{X}_{52|50} = 0.5\hat{X}_{51|50}$ (since $E[W_{52}] = 0$ and $E[W_{51}] = 0$)
-   $h = 3$: $\hat{X}_{53|50} = 0.5\hat{X}_{52|50}$

```{r}
xhat_51 <- 0.5 * xt[n] + 0.5 * wn_hat[n]
xhat_52 <- 0.5 * xhat_51
xhat_53 <- 0.5 * xhat_52
cat("X_50 =", xt[n], "\n")
cat("W_hat_50 =", wn_hat[n], "\n")
cat("X_hat_51 =", xhat_51, "\n")
cat("X_hat_52 =", xhat_52, "\n")
cat("X_hat_53 =", xhat_53, "\n")
```

Using $X_{50} \approx 0.986$ and $\hat{W}_{50} \approx 0.281$:

-   $\hat{X}_{51|50} = 0.5(0.986) + 0.5(0.281) \approx 0.634$
-   $\hat{X}_{52|50} = 0.5(0.634) \approx 0.317$
-   $\hat{X}_{53|50} = 0.5(0.317) \approx 0.158$

The forecasts decay geometrically toward 0 (the process mean) at rate $0.5^h$, since for $h \geq 2$ the MA component no longer contributes and the forecast follows the pure AR(1) recursion.

## 6. ACF, PACF, and BLPs

Let $(X_t)$ be a mean zero stationary process with $\gamma_X(0) = 2$, $\gamma_X(1) = 1.4$, $\gamma_X(2) = 0.6$, $\gamma_X(3) = 0.4$, $\gamma_X(4) = 0.2$.

The autocorrelation function is $\rho(h) = \gamma(h)/\gamma(0)$:

```{r}
gamma_x <- c(2, 1.4, 0.6, 0.4, 0.2)
rho <- gamma_x / gamma_x[1]
names(rho) <- paste0("rho(", 0:4, ")")
rho
```

a.  **Can $(X_t)$ be an MA(2) process?**

No. For an MA(2) process, $\gamma(h) = 0$ for all $|h| > 2$. Since $\gamma(3) = 0.4 \neq 0$, $(X_t)$ cannot be MA(2).

b.  **Can $(X_t)$ be an AR(1) process?**

For AR(1), $\gamma(h) = \phi^h \gamma(0)$. From $\gamma(1)/\gamma(0) = 0.7$, we would need $\phi = 0.7$. Then $\gamma(2)$ should equal $\phi^2\gamma(0) = 0.49 \times 2 = 0.98$. But $\gamma(2) = 0.6 \neq 0.98$, so $(X_t)$ **cannot** be AR(1).

c.  **BLP of $X_4$ given $X_3 = 2$:**

For a mean zero process with a single predictor, $\hat{X}_4 = \frac{\text{Cov}(X_4, X_3)}{\text{Var}(X_3)} X_3 = \frac{\gamma(1)}{\gamma(0)} X_3$.

```{r}
xhat_4c <- (gamma_x[2] / gamma_x[1]) * 2
cat("BLP of X_4 given X_3 = 2:", xhat_4c, "\n")
```

d.  **Variance of $X_4 - \hat{X}_4$** (from part c):

$$
\text{Var}(X_4 - \hat{X}_4) = \gamma(0) - \frac{\gamma(1)^2}{\gamma(0)} = 2 - \frac{1.96}{2} = 1.02
$$

```{r}
var_4c <- gamma_x[1] - gamma_x[2]^2 / gamma_x[1]
cat("Variance:", var_4c, "\n")
```

e.  **BLP of $X_4$ given $X_2 = 2$:**

$$
\hat{X}_4 = \frac{\text{Cov}(X_4, X_2)}{\text{Var}(X_2)} X_2 = \frac{\gamma(2)}{\gamma(0)} X_2
$$

```{r}
xhat_4e <- (gamma_x[3] / gamma_x[1]) * 2
cat("BLP of X_4 given X_2 = 2:", xhat_4e, "\n")
```

f.  **Variance of $X_4 - \hat{X}_4$** (from part e):

$$
\text{Var}(X_4 - \hat{X}_4) = \gamma(0) - \frac{\gamma(2)^2}{\gamma(0)} = 2 - \frac{0.36}{2} = 1.82
$$

```{r}
var_4f <- gamma_x[1] - gamma_x[3]^2 / gamma_x[1]
cat("Variance:", var_4f, "\n")
```

g.  **PACF $\alpha_X(1)$:**

The PACF at lag 1 equals the ACF at lag 1: $$
\alpha_X(1) = \rho(1) = \frac{\gamma(1)}{\gamma(0)} = \frac{1.4}{2} = 0.7.
$$

h.  **PACF $\alpha_X(3)$:**

$\alpha_X(3)$ is the last coefficient $\phi_3$ when solving the Yule-Walker system $\Gamma_3 \boldsymbol{\phi} = \boldsymbol{\rho}_{1:3}$ for an AR(3) model:

```{r}
Gamma3 <- matrix(c(1, rho[2], rho[3],
                    rho[2], 1, rho[2],
                    rho[3], rho[2], 1), nrow = 3, byrow = TRUE)
phi3 <- solve(Gamma3, rho[2:4])
names(phi3) <- paste0("phi_", 1:3)
phi3
cat("alpha_X(3) =", phi3[3], "\n")
```

From the Yule-Walker solution, the AR(3) coefficients are $\hat\phi_1 \approx 1.107$, $\hat\phi_2 \approx -0.750$, $\hat\phi_3 \approx 0.393$. The PACF at lag 3 is $\alpha_X(3) = \hat\phi_3 \approx 0.393$, which is nonzero. This further confirms that the process is not AR(1) or AR(2), since for an AR($p$) process, $\alpha_X(h) = 0$ for all $h > p$.
