---
title: "ST5209/X Assignment 5"
format: pdf
editor: visual
author: "Jiao Dian"
execute:
  warning: false
---

## Submission

1.  Render the document to get a .pdf printout.
2.  Submit both the .qmd and .pdf files to Canvas.

```{r}
#| message: FALSE
#| echo: FALSE
#| output: false
Sys.setlocale("LC_ALL", "English")
options(repos = c(CRAN = "https://cloud.r-project.org"))
library(tidyverse)
library(fpp3)
```

## 1. Modeling with ARIMA (Q9.7 in Hyndman & Athanasopoulos)

```{r}
aus_airpassengers |> autoplot(Passengers) +
  labs(title = "Australian air passengers", y = "Passengers (millions)")
```

a.  **Fit an ARIMA model with `ARIMA()`.**

```{r}
fit_auto <- aus_airpassengers |>
  model(arima = ARIMA(Passengers ~ 0 + pdq(0, 2, 1)))
report(fit_auto)
```

Selected: **ARIMA(0, 2, 1)** (no constant), $\hat\theta_1 = -0.8963$ (s.e. 0.0594), $\hat\sigma^2 = 4.308$, AICc = 198.32.

```{r}
fit_auto |> gg_tsresiduals()
fit_auto |> augment() |> features(.innov, ljung_box, lag = 10, dof = 1)
```

All ACF spikes lie inside the bands and the Ljung–Box test gives $Q^* = 6.70$, $p = 0.669$ — residuals are consistent with white noise.

```{r}
fit_auto |>
  forecast(h = 10) |>
  autoplot(aus_airpassengers) +
  labs(title = "ARIMA(0,2,1) 10-year forecast", y = "Passengers (millions)")
```

Forecasts 2017–2026: $74.8, 77.0, 79.2, 81.3, 83.5, 85.7, 87.9, 90.1, 92.3, 94.5$ — nearly linear with slope $\approx 2.2$ per year.

b.  **Backshift operator form.**

For ARIMA(0, 2, 1), the model is $(1-B)^2 X_t = (1 + \theta_1 B) W_t$. Substituting $\hat\theta_1 = -0.8963$,

$$
(1 - B)^2 X_t = (1 - 0.8963\,B)\, W_t, \qquad W_t \sim WN(0, 4.308).
$$

c.  **ARIMA(0, 1, 0) with drift.**

```{r}
fit_c <- aus_airpassengers |>
  model(arima = ARIMA(Passengers ~ 1 + pdq(0, 1, 0)))
report(fit_c)
fit_c |>
  forecast(h = 10) |>
  autoplot(aus_airpassengers) +
  labs(title = "ARIMA(0,1,0) with drift, 10-year forecast")
```

Random walk with drift $\hat c = 1.4191$ (AICc = 200.59). Forecasts $74.0, 75.4, \dots, 86.8$ — a straight line with slope 1.42, smaller than part (a)'s $\approx 2.2$. Since ARIMA(0,2,1) implies $\Delta X_t$ follows an MA(1), its forecast extrapolates the *recent* change in slope rather than a fixed drift, better matching the steepening growth near the end of the series.

d.  **ARIMA(2, 1, 2) with drift.**

```{r}
fit_d <- aus_airpassengers |>
  model(arima = ARIMA(Passengers ~ 1 + pdq(2, 1, 2)))
report(fit_d)
fit_d |>
  forecast(h = 10) |>
  autoplot(aus_airpassengers) +
  labs(title = "ARIMA(2,1,2) with drift, 10-year forecast")
```

AICc = 206.61 (worse than part (a)'s 198.32). Forecasts $73.2, 75.6, 77.1, 77.7, 79.5, \dots, 86.6$ show a mild wavy pattern from the complex AR roots, tracking slightly below the ARIMA(0,2,1) forecasts.

**Removing the constant.**

```{r}
fit_d2 <- tryCatch(
  aus_airpassengers |>
    model(arima = ARIMA(Passengers ~ 0 + pdq(2, 1, 2))),
  error = function(e) e
)
fit_d2
```

Without the drift, `ARIMA()` returns `NULL` with the message *"non-stationary AR part from CSS"*: the CSS optimiser cannot find a stationary AR, so estimation fails.

e.  **ARIMA(0, 2, 1) with a constant.**

```{r}
fit_e <- aus_airpassengers |>
  model(arima = ARIMA(Passengers ~ 1 + pdq(0, 2, 1)))
report(fit_e)
fit_e |>
  forecast(h = 10) |>
  autoplot(aus_airpassengers) +
  labs(title = "ARIMA(0,2,1) with constant, 10-year forecast")
```

`ARIMA()` warns of a polynomial trend. With $d=2$ and a non-zero constant $c$, $X_t$ carries a quadratic term $\tfrac{c}{2}t^2$. Forecasts $75.4, 78.2, \dots, 103.0$ curve upward (yearly increments grow from 2.8 to 3.3). This quadratic extrapolation is unrealistic; prefer the model in (a).

## 2. Seasonal ARIMA

a.  **Load the data.**

```{r}
diabetes <- read_rds("../_data/cleaned/diabetes.rds")
head(diabetes)
```

b.  **Build `Y`: log + seasonal difference.**

```{r}
dbt <- diabetes |> mutate(Y = difference(log(Cost), lag = 12))
```

c.  **KPSS test on `Y`.**

```{r}
dbt |> features(Y, unitroot_kpss)
```

KPSS stat $= 0.209$, p-value $= 0.1$ (capped at 0.1). We fail to reject stationarity, so `Y` is consistent with being stationary.

d.  **Time plot of `log(Cost)`.**

```{r}
dbt |> autoplot(log(Cost)) +
  labs(title = "log(Cost)")
```

`log(Cost)` has a clear upward trend plus annual seasonality. If $\log(\text{Cost}_t) \approx \alpha + \beta t + s_t + u_t$ with period-12 $s_t$, then $Y_t = 12\beta + (u_t - u_{t-12})$: the trend collapses to the constant $12\beta$ and the seasonal component cancels.

e.  **Fit an ARIMA model to `log(Cost)`.**

```{r}
fit2 <- dbt |> model(arima = ARIMA(log(Cost)))
report(fit2)
```

Fitted model: **ARIMA(2,0,2)(1,1,1)$_{12}$ with drift** ($\hat\sigma^2 = 0.00360$, AICc = $-515.24$).

f.  **Number of fitted parameters.**

```{r}
tidy(fit2)
```

**7 fitted coefficients** (`ar1, ar2, ma1, ma2, sar1, sma1, constant`). Counting $\sigma^2$, $k = 8$, consistent with $\mathrm{AIC} + 2\log L = -516.02 + 532.02 = 16 = 2k$.

## 3. Model selection

a.  **Null hypothesis and assumptions of the ADF test.**

$H_0$: the series has a **unit root** ($I(1)$); $H_1$: stationary (possibly around a trend). The ADF regresses $\Delta X_t$ on $X_{t-1}$ and $p$ lagged differences. Assumptions: linear AR data-generating process with mean-zero, finite-variance, (approximately) white-noise innovations; $p$ large enough to whiten residuals; and the included deterministic terms (constant / trend) match those of the data under $H_1$.

b.  **Can ADF and KPSS both have large p-values?**

Yes. ADF and KPSS have opposite nulls (unit root vs stationarity), so large p-values mean *failing to reject* their respective nulls. In short or borderline samples (e.g. $\phi$ close to 1), both tests can have low power, so neither rejects. The conclusion is then that the data are inconclusive — use differencing + re-testing, plots, or domain knowledge.

c.  **AIC and AICc penalties for the Q2 model.**

With $k = 8$ and $n = 204$:

```{r}
n <- nrow(diabetes); k <- 8
c(AIC_penalty = 2 * k,
  AICc_penalty = 2 * k + 2 * k * (k + 1) / (n - k - 1))
```

$\text{AIC penalty} = 2k = \mathbf{16}$; $\text{AICc penalty} = 16 + 144/195 \approx \mathbf{16.738}$.

d.  **Fit an ETS model and compare.**

```{r}
fit_both <- diabetes |>
  model(
    arima = ARIMA(log(Cost)),
    ets   = ETS(Cost)
  )
glance(fit_both) |> select(.model, log_lik, AIC, AICc, BIC)
```

ETS selects **ETS(M, A, M)**. Log-likelihoods and AICc:

| Model | log L | AICc |
|----|----|----|
| ARIMA on $\log$(Cost) | $+266.01$ | $-515.24$ |
| ETS(M, A, M) on Cost | $-416.97$ | $871.23$ |

e.  **Can we compare them via log-likelihood or AICc?**

**No.** The two are on different scales ($\log$(Cost) vs Cost), and the Jacobian shifts the ARIMA log-likelihood by $-\sum_t \log(\text{Cost}_t)$. To compare, fit both on the same scale, adjust the ARIMA log-likelihood, or use time-series cross-validation.

## 4. Dynamic regression (adapted from Q10.4 in Hyndman & Athanasopoulos)

a.  **Inflation adjustment.**

```{r}
acc <- aus_accommodation |>
  mutate(AdjTakings = Takings / CPI * 100)
```

b.  **Fit a dynamic regression model for each state.**

```{r}
knot <- yearquarter("2008 Q1")
fits <- acc |>
  model(dr = ARIMA(AdjTakings ~ season() + trend(knots = knot)))
fits
```

Each row shows the ARIMA error structure chosen per state.

c.  **Victoria.**

```{r}
vic <- fits |> filter(State == "Victoria")
report(vic)
```

Victoria: linear regression with **ARIMA(1,0,0)(0,0,1)$_4$** errors. Intercept $= 264.4$; seasonal dummies (Q1 reference) year2 $= -57.9$, year3 $= -36.0$, year4 $= -12.9$ (all significant); pre-knot slope $= 3.20$ (significant); post-knot increment $= 0.389$ ($p = 0.69$, not significant); AR1 $= 0.660$, SMA1 $= 0.376$ ($p = 0.003$).

**Constant seasonality?** The significant SMA1 shows that **residual seasonal dependence remains after the fixed dummies** — the dummies alone do not fully capture the seasonal structure, and the SMA1 term absorbs the leftover seasonal correlation.

d.  **Residual diagnostics for Victoria.**

```{r}
vic |> gg_tsresiduals()
vic |> augment() |> features(.innov, ljung_box, lag = 8, dof = 0)
```

No ACF spike falls outside the bands; Ljung–Box at lag 8 gives $Q^* = 4.90$, $p = 0.769$. Residuals are consistent with white noise.

e.  **Forecast CPI, then forecast Victoria takings to the end of 2017.**

The data ends in 2016 Q2, so we need $h = 6$ quarters (2016 Q3 through 2017 Q4).

```{r}
# Forecast CPI separately with its own ARIMA model
cpi_ts <- acc |>
  filter(State == "Victoria") |>
  select(Date, CPI) |>
  as_tsibble(index = Date)
fit_cpi <- cpi_ts |> model(cpi_arima = ARIMA(CPI))
report(fit_cpi)
fcst_cpi <- fit_cpi |> forecast(h = 6)

# Supply forecasted CPI as new_data to the Victoria dynamic regression
new_data_vic <- new_data(acc |> filter(State == "Victoria"), 6) |>
  mutate(CPI = fcst_cpi$.mean)

fcst_adj <- vic |> forecast(new_data = new_data_vic)

takings_fcst <- as_tibble(fcst_adj) |>
  select(Date, AdjTakings_hat = .mean) |>
  mutate(CPI = new_data_vic$CPI,
         Takings_hat = AdjTakings_hat * CPI / 100)
takings_fcst
```

CPI: **ARIMA(0,1,0) with drift** ($\hat c = 0.570$ per quarter). Re-inflating via $\widehat{\text{Takings}} = \widehat{\text{AdjTakings}} \cdot \widehat{\text{CPI}} / 100$:

| Quarter | $\widehat{\text{AdjTakings}}$ | $\widehat{\text{CPI}}$ | $\widehat{\text{Takings}}$ (A\$M) |
|-----------------|-----------------|-----------------|---------------------|
| 2016 Q3 | 473.99 | 109.17 | **517.45** |
| 2016 Q4 | 511.67 | 109.74 | **561.50** |
| 2017 Q1 | 523.98 | 110.31 | **578.01** |
| 2017 Q2 | 463.23 | 110.88 | **513.62** |
| 2017 Q3 | 491.15 | 111.45 | **547.39** |
| 2017 Q4 | 519.47 | 112.02 | **581.90** |

```{r}
acc |>
  filter(State == "Victoria") |>
  ggplot(aes(x = Date, y = Takings)) +
  geom_line() +
  geom_line(data = takings_fcst, aes(y = Takings_hat), colour = "red") +
  labs(title = "Victoria total takings — history and 6-quarter forecast",
       y = "Takings (A$M)")
```

Forecast preserves the Q1-peak / Q2-trough pattern and continues the pre-2008 linear rise.

f.  **Sources of uncertainty not reflected in the intervals.** The intervals propagate only the Victoria ARIMA innovation variance. They ignore: (i) CPI forecast uncertainty (point forecasts were plugged in as regressors); (ii) parameter estimation uncertainty in the Victoria model; (iii) model/specification uncertainty (knot at 2008 Q1, piecewise-linear trend, deterministic dummies, ARIMA(1,0,0)(0,0,1)$_4$ errors).

## 5. State space models

Consider $$
T_t = T_{t-1} + \epsilon_t, \quad
S_t = 0.8\, S_{t-4} + \eta_t, \quad
Y_t = T_t + S_t + V_t,
$$ with $\epsilon_t \sim WN(0, \sigma_1^2), \eta_t \sim WN(0, \sigma_2^2), V_t \sim WN(0, \sigma_3^2)$, mutually uncorrelated.

a.  **Linear Gaussian state space form.**

Since $S_t$ depends on $S_{t-4}$, keep three lags in the state: $$\mathbf{X}_t = (T_t, S_t, S_{t-1}, S_{t-2}, S_{t-3})^\top.$$ Then

$$
\mathbf{X}_t = F\,\mathbf{X}_{t-1} + \mathbf{w}_t, \qquad
Y_t = H\,\mathbf{X}_t + V_t,
$$

with $$
F = \begin{pmatrix}
1 & 0 & 0 & 0 & 0 \\
0 & 0 & 0 & 0 & 0.8 \\
0 & 1 & 0 & 0 & 0 \\
0 & 0 & 1 & 0 & 0 \\
0 & 0 & 0 & 1 & 0
\end{pmatrix}, \quad
\mathbf{w}_t = \begin{pmatrix} \epsilon_t \\ \eta_t \\ 0 \\ 0 \\ 0 \end{pmatrix}, \quad
Q = \mathrm{Cov}(\mathbf{w}_t) = \mathrm{diag}(\sigma_1^2,\, \sigma_2^2,\, 0,\, 0,\, 0),
$$

and $$
H = \begin{pmatrix} 1 & 1 & 0 & 0 & 0 \end{pmatrix}, \qquad R = \mathrm{Var}(V_t) = \sigma_3^2.
$$

Row 1 gives $T_t = T_{t-1} + \epsilon_t$, row 2 gives $S_t = 0.8 S_{t-4} + \eta_t$ (the 5th entry of $\mathbf{X}_{t-1}$ is $S_{t-4}$), rows 3–5 shift the seasonal lags, and $H\mathbf{X}_t + V_t = T_t + S_t + V_t = Y_t$.

b.  **Does $\lim_{t \to \infty}\mathrm{Var}(S_t - \hat S_{t|n})$ exist?**

With $n$ fixed and $t \to \infty$, $\hat S_{t|n}$ is a long-horizon forecast. Since $|0.8| < 1$, $S_t$ is stationary with $\gamma_S(0) = \sigma_2^2/(1-0.64) = \sigma_2^2/0.36$. As $t \to \infty$, $\hat S_{t|n} \to 0$, so the limit **exists** and equals

$$\lim_{t \to \infty}\mathrm{Var}(S_t - \hat S_{t|n}) = \frac{\sigma_2^2}{1 - 0.8^2} = \frac{25\sigma_2^2}{9}.$$

c.  **Does $\lim_{t \to \infty}\mathrm{Var}(T_t - \hat T_{t|n})$ exist?**

$T_t$ is a random walk, and for $t > n$, $\hat T_{t|n} = \hat T_{n|n}$. Hence

$$\mathrm{Var}(T_t - \hat T_{t|n}) = \mathrm{Var}(T_n - \hat T_{n|n}) + (t-n)\sigma_1^2 \to \infty.$$

The limit **does not exist** — a random walk's forecast variance grows without bound.
